???A quantitative framework for investigating the reliability of network construction

### 2017-09-27 manuscript/CI_interactions.tex
moving on to the Bayesian approach section

A simple example will illustrate the problem: assume that we want to infer the probability of an interaction between two species, A and B. Now assume that in reality, interaction between A and B is completely impossible (i.e. the true (P(L|co-occurrence of A and B = 0). Given no added detection error in observing the incidence of the interaction, then a single observation of species co-occurrence will reveal (very) little regarding the probability of the interaction – the confidence limits for one observation essentially spans from 0 to 1 (see following section). For 5 observations, this confidence limit will still extends to 0.5. Only with thirty observations will it shrink to 0.1. Thus, adding (many) more observations is certainly useful in controlling uncertainty due to stochasticity inherent in the process itself.

@@

The pattern of Fig.~\ref{upper_limits} should be compared to the frequency of co-occurrence of taxa in the Salix-galler data set. Here, most(TO COMPUTE) pairs of species are never found co-occurring, and for species found co-occurring the number of co-occurrences was generally low (Fig from AC and KW). Among the species which did sometimes interact when co-occurring, the incidence of interaction was actually low(TO COMPUTE) (freq histogram Fig from AC and KW). Thus, even in the most extensive data set that we could think of, limited sampling is clearly a major source of uncertainty.

The obvious rule of thumb derived from the above example will be “sample more”, but we note that there are limits to the utility of this motto. Sampling more will clearly reduce uncertainty regarding the upper bound of the probability of interaction, and it will also increase the probability of detecting unlikely interactions (i.e. interactions occurring with a low probability). Yet, since the probability of observing the co-occurrence of two species will always be higher than the probability of observing their interaction (since the probability of interaction is conditional on both interaction partners being present), we will accumulate observations of co-occurrences quicker than we will accumulate observations of interactions. Thus, the more we sample, the more holes will appear in our interaction matrix, which may seem like a counter-intuitive outcome.

As a solution towards deciding whether unobserved interactions were undetected due to sampling, or whether they truly do not occur, Weinstein et al. (2017) used daily repeated sampling rounds to estimate the daily probability of detecting a hummingbird interaction, and to thereby model the observation and process mechanisms (Weinstein, B. G., \& Graham, C. H. (2017). On comparing traits and abundance for predicting species interactions with imperfect detection. Food Webs 11; 17-25: https://doi.org/10.1016/j.fooweb.2017.05.002). While conceptually attractive, this approach is unsuitable for interactions occurring over longer time scales (like associations between hosts and parasitoids with a single generation per year), or very rare interactions (since no matter how frequently and extensively we sample, there will still be some rare interactions lurking around). What is worse, the problem persists that if a given interaction is not observed on a given day, then it may still be either because it was impossible that day ((P(L|co-occurrence of A and B)=0), because interaction between the two did occur but was not observed, or from any combination of the two. Thus, from a conceptual perspective, this approach will fail to satisfactorily separating between all sources of uncertainty. An interaction not observed on day t may have gone unobserved either because there were constraints on the process (i.e. the probability was really zero), because of chance variation in the realized incidence of an event associated with a finite probability. Most importantly, if during a series of days, two species are simply never observed co-occurring, then we still have not learnt a thing about their probability of interacting should they do so. In other words, if the two interaction partners do not co-occur, we will still gain no information.

An added complication is that not all sourced of uncertainty are proportional to sample size. To score an interaction between A and B, we will need to identify both partners correctly (a non-trivial problem in many food webs; e.g. Kaartinen et al. 2011 Ecol Entomol; Roslin and Majaneva 2016) and be able to resolve all interactions with a similar likelihood. For both molecular (ref ref) and rearing techniques, given types of interactions may go unnoticed due to technical challenges.

The bottom line is that separating between different sources of uncertainty is difficult indeed. In the best of all possible worlds, we might then use an experimental approach to estimate detection uncertainty: we might create replicate sets of cases where the interaction is really there, then evaluating them and examining the fraction of cases scoring positive. Yet, as anyone attempting these experiments will know, most interaction partners are proving notoriously unwilling to collaborate. Gallers refuse to lay eggs on their host plants (ref ref), parasitoids to attack their hosts when experimentally introduced (Gariepy 2017; ref ref). If we introduce only one of the components (like in the cases where we expose a set of hosts to parasitism; ref ref; Gripenberg et al. xxx) then we are already confounding multiple levels of uncertainty in the outcome – that of interaction uncertainty (whether the species will interact once both there), process uncertainty (whether the interaction will occur if and since the probability of interaction is below one), and detection uncertainty (i.e. whether we will observed the interaction once realized). To perform these experiments in a comprehensive and reliable manner, we will then need to expose a large number of interactions in a comprehensive experiment across space. This type of “cafeteria experiments in space” is not only unfeasible, but also undermining our original objective: If the local occurrence of interactions is uninformative regarding the underlying probabilities of occurrence, then we should be doing something else than wasting a monumental effort on collecting local interaction matrices. <deep sigh and forget that section? >

As a tentative solution to disentangling observation and process mechanisms, we propose that some insights regarding the detectability of interactions between species not found co-occurring may be gained from data on other species, for which it is known that the interaction is feasible. If we use this information obtain the joint probability of detection and process (e.g. to glean information regarding at how many sites the species are found interacting across the entire dataset), and assuming that most of this pattern is due to variation in detection rate, then insights into the incidence of interactions between species may be gleaned from the accumulation rate of positive interactions at sites were interaction is possible (i.e. with at least one observation of a positive interaction) [this reasoning to be further formalized]

>>>>

  A simple example will illustrate the problem: assume that we want to infer the probability of an interaction between two species, A and B. Now assume that in reality, interaction between A and B is completely impossible (i.e. the true $\lambda$= 0). Given no added detection error in observing the incidence of the interaction, then a single observation of species co-occurrence will reveal (very) little regarding the probability of the interaction – the confidence limits for one observation essentially spans from 0 to 1 (see following section). For 5 observations, this confidence limit will still extends to 0.5. Only with thirty observations will it shrink to 0.1. Thus, adding (many) more observations is certainly useful in controlling uncertainty due to stochasticity inherent in the process itself.

@@

  The pattern of Fig.~\ref{upper_limits} should be compared to the frequency of co-occurrence of taxa in the Salix-galler data set. Here, most (3986 out of 4992) pairs of species are never found co-occurring, and for species found co-occurring the number of co-occurrences was generally low (Fig.~\ref{histograms}A). Among the species which did sometimes interact when co-occurring, the incidence of interaction was actually low. Only 141 pairs of species were observed interacting: of these, 26 were observed interacting only once and 48.2\% were observed interacting five or fewer times (Fig.~\ref{histograms}B). Thus, even in the most extensive data set that we could think of, limited sampling is clearly a major source of uncertainty.

  \begin{figure}
    \caption{\textbf{A)} Most pairs of \emph{Salix} and galler species were never observed co-occurring despite the high levels of replication in our example dataset. For those that were observed together at least once, the number of observed co-occurances was generally small (\textless10). Here we show a histogram of the number of pairs of species observed co-occurring at least once. 3986 pairs of species were never observed co-occurring: this is represented by the upwards triangle at 0. \textbf{B)} Most pairs of species that were observed at the same site were never observed interacting. Here we show a histogram of the number of observed interactions within pairs of co-occurring species.}
    \label{histograms}
    \includegraphics*[width=.8\textwidth]{figures/Salix_Galler_histogram.eps}
    \end{figure}

  The obvious rule of thumb derived from the above example is “sample more”, but we note that there are limits to the utility of this motto. Sampling more will clearly reduce uncertainty regarding the upper bound of the probability of interaction, and it will also increase the probability of detecting unlikely interactions (i.e. interactions occurring with a low probability). Yet, since the probability of observing the co-occurrence of two species will always be higher than the probability of observing their interaction (since the probability of interaction is conditional on both interaction partners being present), we will accumulate observations of co-occurrences faster than we will accumulate observations of interactions. Thus, the more we sample, the more zeros will appear in our interaction matrix, which may seem like a counter-intuitive outcome.

  As a solution towards deciding whether unobserved interactions were undetected due to sampling, or whether they truly do not occur,~\citet{Weinstein2017} used daily repeated sampling rounds to estimate the daily probability of detecting a hummingbird interaction, and to thereby model the observation and process mechanisms~\citep{Weinstein2017}. While conceptually attractive, this approach is unsuitable for interactions occurring over longer time scales (e.g., associations between hosts and parasitoids with a single generation per year), or very rare interactions (since no matter how frequently and extensively we sample, there will still be some interactions lurking around). What is worse, the problem persists that if a given interaction is not observed on a given day, this could be either because it was impossible that day despite being otherwise feasible (P($X|L=1$=0), because interaction between the two did occur but could not be observed (P($D|X$=1,L=1)=0), or from any combination of the two. From a conceptual perspective, this approach will thus fail to satisfactorily distinguish between sources of uncertainty. An interaction not observed on day $t$ may have gone undetected either because there were constraints on the process (i.e. the probability $\lambda$ was really zero) or because of chance variation in the realized incidence of an event associated with a finite probability. Most importantly, if two species are never observed co-occurring during several days of sampling then we have learned nothing about their probability of interacting if they should ever co-occur. In other words, there is no information without co-occurance.

  An added complication is that not all sources of uncertainty are proportional to sample size. To record an interaction between A and B, we need to identify both partners correctly (a non-trivial problem in many food webs; e.g.~\citet{Kaartinen2011,Roslin2016}) and be able to resolve all interactions with a similar likelihood. For both molecular~\citep{} and rearing~\citep{} techniques, certain types of interactions may go unnoticed due to technical challenges. This can bias the set of recorded interactions.

  The bottom line is that separating different sources of uncertainty is difficult indeed. In the best of all possible worlds, we might use an experimental approach to estimate detection uncertainty by creating replicate cases where the interaction really occurs and then evaluating the fraction of cases where the interaction is detected. Yet, as anyone attempting these experiments will know, most interaction partners are notoriously unwilling to collaborate. Gallers refuse to lay eggs on their host plants in captivity (ref ref) and parasitoids decline to attack their hosts when experimentally introduced (Gariepy 2017; ref ref). If we introduce only one of the components (as in the cases where we expose a set of hosts to parasitism; ref ref; Gripenberg et al. xxx) then we are already confounding multiple levels of uncertainty since it is not certain whether the interaction will occur or whether we can detect it. To perform these experiments in a comprehensive and reliable manner, we will then need to replicate a large number of interactions in a comprehensive experiment across space. This type of “cafeteria experiments in space” is not only unfeasible, but also undermines our original objective: If the local occurrence of interactions is uninformative regarding the underlying probabilities of occurrence, then is not likely to be worthwhile collecting these local interaction matrices in the first place. <deep sigh and forget that section? >

  As an alternative to abandoning empirical networks or continuing to ignore the uncertainty inherent in undetected observations, we propose that some insight regarding the detectability of interactions between species not found co-occurring may be gained from data on other species between which interactions are known to be feasible. If we use this information obtain the joint probability of detection and process (e.g. to glean information regarding at how many sites the species are found interacting across the entire dataset), and assuming that most of this pattern is due to variation in detection rate, then insights into the incidence of interactions between species may be gleaned from the accumulation rate of positive interactions at sites were interaction is possible (i.e. with at least one observation of a positive interaction) [this reasoning to be further formalized]
